---
title: "Assignment 2: Coding Basics"
<<<<<<< HEAD
author: "Max Hermanson "
=======
author: "Student Name"
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

<<<<<<< HEAD
4. Insert comments in your code to describe what you are doing. 

```{r}
#1. 
sequence1  <- seq(1,100,4)
sequence1 

#2. 
mean_sequence1 <- mean(sequence1)
mean_sequence1
#3. 
median_sequence1 <- median(sequence1)
median_sequence1

mean_vs_median <- function(mean_sequence1){
  if(median_sequence1 == mean_sequence1){
    print("true")}
  else {print("false")}
}
mean_vs_median
=======
4. Insert comments in your code to describe what you are doing.

```{r}
#1. 

#2. 

#3. 

>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}

<<<<<<< HEAD


name_student <- c("Max", "Maria", "Joe", "Timor")
test_scores <- c(50, 70, 90, 59)
class(name_student)
class(test_scores)
pass_fail <- function(test_scores){
  if(test_scores<50){test_scores<- TRUE}
  else{test_scores<- FALSE}
}
pass_fail
```

```{r 2}


name_student <- c("Max", "Maria", "Joe", "Timor")
test_scores <- c(50, 70, 90, 59)
pass_fail2 <- function(test_scores){
  if(test_scores<50){x2<- "TRUE"}
  else{x2<- "FALSE"}
  return(list(x2=x2))}

output_testscore<-pass_fail2(test_scores)
output_testscore
```

```{r DF test}

student_df<- data.frame(name_student, test_scores, pass_fail2)
student_df
```


```{r Take2}
Name <- c("Max", "Maria", "Joe", "Timor") #character
Test_Score <- c(40, 70, 90, 59) #numeric
Passing<- c(FALSE, TRUE, TRUE, TRUE) #logical
class(grade3)

student_ID_Grade_df<- data.frame(Name, Test_Score, Passing)
student_ID_Grade_df

```

```{r Take 3 ifelse}
Name <- c("Max", "Maria", "Joe", "Timor") #character
Test_Score <- c(40, 70, 90, 59) 

pass_fail <- function(x){
  ifelse(x>50, print("True"), print("False"))
}
tester <- pass_fail(Test_Score)
tester
=======
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
