---
title: "Assignment 3: Data Exploration"
<<<<<<< HEAD
author: "Max Hermanson     "
=======
author: "Student Name"
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 
<<<<<<< HEAD
fasdfas
=======

>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 
```{r,  message = FALSE}

<<<<<<< HEAD

getwd()
setwd("/Users/mothership/Desktop/EDA_21/Environmental_Data_Analytics_2021_2/Assignments")
getwd()
neonic.df <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
litter.df <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv")

library(ggplot2)
=======
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

<<<<<<< HEAD
> Answer: research has shown that neonicotinoid pesticides, which are widely used on farms in the US, may adversely affect pollinators and other important insects. Such effects could have wide-reaching economic, social, and biological implications, and should thus be studied. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: litter and woody debris can be a good determinant of forest productivity, local soil fertility, and nutrient cycling, and is an important source of habitat for a variety of organisms. 
=======
> Answer: 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: 
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
<<<<<<< HEAD
* Tower and ground traps collect the litter/debris
* These traps are randomly placed within 400m^2 plots. Randomness is maximized by using different randomness placement strategies for different forest cover types  
* The ground traps are sampled once every year, whereas the tower (air) traps are sampled much more frequently, but this frequency depends on site type. 
=======
*
*
*
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
<<<<<<< HEAD
dim(neonic.df)
=======

>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

6. Using the `summary` function on the "Effects" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
<<<<<<< HEAD
summary(neonic.df$Effect)
head(neonic.df)

```

> Answer: Mortality, Population, and Behavior have the highest numbers of recorded observations. Mortality and population are both metrics of abundance for insect populations of interest, which can be good indicators of fitness. If a strong correlation exists between neonic use and population changes, this could be very informative in neonic policy. Insights into behavior could be both an indirect measure of fitness and potential changes in pollination activity, which could be extremely important given how reliant many foodwebs are on pollination. 
=======

```

> Answer: 
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
<<<<<<< HEAD
summary(neonic.df$Species.Common.Name)

```

> Answer: honey bee, parasitic wasp, buff tailed bumblebee, carniolan honey bee, bumble bee, italian honeybee. Five out of six of these are bees/pollinators. They are likely of interest because they are major insect pollinators, which may make them most susceptible to neonic exposure upon pollinating agricultural crops. It is possible the parasitic wasp populations are correlated with bee populations due to a parasitic relationship. 
=======

```

> Answer: 
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
<<<<<<< HEAD
class(neonic.df$Conc.1..Author.)

```

> Answer: conc.1..Author is in the factor data format. Perhaps because dosage concentrations were kept constant for each type of neonic studied, and were thus stored as categorical data instead of numeric data. 
=======

```

> Answer:
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}

<<<<<<< HEAD
ggplot(neonic.df)+
  geom_freqpoly(aes(x=Publication.Year))

=======
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
<<<<<<< HEAD
ggplot(neonic.df)+
  geom_freqpoly(aes(x=Publication.Year, color = Test.Location))
=======

>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

Interpret this graph. What are the most common test locations, and do they differ over time?

<<<<<<< HEAD
> Answer: The most common test locations are in the lab and field. Field studies seem to have dominated the number of studies in the 90's, but lab studies began to take over from 2001 onwards. It appears research on neonics heavily declined between 2015 and 2016. 
=======
> Answer: 
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}

<<<<<<< HEAD
ggplot(neonic.df)+
  geom_bar(aes(x=Endpoint))
```

> Answer: LOEL and NOEL are the two most common endpoints. 

>LOEL endpoints are defined as: "Lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls (LOEAL/LOEC)" - metadata pdf 
>NOEL endpoints are defined as: "No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test (NOEAL/NOEC)". 
=======
```

> Answer: 
>
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
<<<<<<< HEAD
class(litter.df$collectDate)
litter.df$collectDate[1:10]
litter.df$collectDate <- as.Date(litter.df$collectDate, format = "%Y-%m-%d")
litter.df$collectDate
class(litter.df$collectDate)
=======

>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
<<<<<<< HEAD
Niwot_ridge_plots <- unique(litter.df$plotID)
Niwot_ridge_summary<-summary(litter.df$plotID)
#litter.df$siteID[1:100]
length(Niwot_ridge_plots)
Niwot_ridge_summary


```

> Answer: 12 plots were sampled at Niwot Ridge. The 'unique' function delineates subgroups within variables, whereas 'provides' general information about each variable as a whole (length, class, mode ). 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(litter.df)+
  geom_bar(aes(x=functionalGroup, color=functionalGroup))
=======

```

> Answer: 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}

>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
<<<<<<< HEAD
ggplot(litter.df)+
  geom_boxplot(aes(x=functionalGroup, y=dryMass))

ggplot(litter.df)+
  geom_violin(aes(x=functionalGroup, y=dryMass), draw_quantiles = c(0.25,0.5,0.75))

ggplot(litter.df)+
  geom_point(aes(x=functionalGroup, y=dryMass))
=======

>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

<<<<<<< HEAD
> Answer: The violin plots look extremely stretched, possibly due to a wide range of values and the presence of outliers. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Mixed litter, Needles, and Twigs/Branches tend to have the highest biomass at these sites. This was determined by examining the boxplots of functionGroup v Litter. A greater proportion of the samples from these three variables weigh significantly more than the other variables, as seen by the interquartile range of their boxplots.  
=======
> Answer: 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: 
>>>>>>> 51cde3a2ab3ab7ade34cb8f398b19630d942ece2
