---
title: "Assignment 5: Data Visualization"
author: "Max Hermanson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 23 at 11:59 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (both the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] and the gathered
[`NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv`] versions) and 
the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
library(cowplot)
library(tidyverse)
library(ggridges)

#2 
getwd()

lake_processed <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv") 
lake_gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
litter.df <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}

maxsTheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

#theme_set(maxsTheme)


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
phosphorus_scatter <- ggplot(lake_processed,aes(x= tp_ug, y=po4))+
         geom_point(aes(color = lakename))+
  maxsTheme+
  xlab("Total Phosphorus")+
  ylab("Phosphate Concentration")+
  geom_smooth(method=lm, color = "black")+
  ylim(0,50)+
  labs(lakename = "Lake Name:") #cant get to work 
print(phosphorus_scatter)  




phosphorus_dot <- ggplot(lake_processed,aes(x= tp_ug, y=po4))+
         geom_dotplot(aes(color = lakename), binaxis = "y", binwidth = 1 , stackdir = "center", position = "dodge", dotsize = 1)+
  maxsTheme+
  xlab("Total Phosphorus")+
  ylab("Phosphate Concentration")+
  ylim(0,100)
print(phosphorus_dot)  

#my choice 
phosphorus_violin <- ggplot(lake_processed,aes(x= tp_ug, y=po4, fill = lakename))+
  geom_violin()+
  ylim(0,50)+
  facet_wrap(vars(lakename), nrow = 1)+
  xlab("Total Phosphorus")+
  ylab("Phosphate Concentration")+
  maxsTheme
print(phosphorus_violin)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
temp_box <- ggplot(lake_processed, aes(x = month, y = temperature_C))+
  geom_boxplot(aes(color = as.factor(month)))+
  xlab("Month")+
  ylab("Temperature C")
  
print(temp_box)
  
TP_box <- ggplot(lake_processed, aes(x = month, y = tp_ug))+
  geom_boxplot(aes(color = as.factor(month)))+
  xlab("Month")+
  ylab("Total Phosphorus")
  
  
print(TP_box)

TN_box <- ggplot(lake_processed, aes(x = month, y = tn_ug))+
  geom_boxplot(aes(color = as.factor(month)))+
  ylim(0, 3000)+
  xlab("Month")+
  ylab("Total Nitrogen")
  

print(TN_box)


combined_cow <- plot_grid(temp_box, TP_box, TN_box, nrow = 3, align = 'h', rel_heights = c(0.75, 1, 1))
print(combined_cow)

ggsave("./Output/A05_cowplot.png", combined_cow, dpi = 300)


facet_wrap(vars(lakename), nrow = 1)+
  xlab("Total Phosphorus")+
  ylab("Phosphate Concentration")+


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
# Dry mass = Y axis, Collect Date = X-Axis? > facet wrap problem: make sure you use vars command 
# Faced grid, more than 1 variable 
# Facet wrap, 
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
