---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Max Hermanson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
library(tidyverse)
#install.packages("agricolae")
library(agricolae)
library(ggridges)
library(lubridate)

lakechem.df <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

#2
maxstheme2 <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(maxstheme2)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: mean lake temperature does not change with depth across lakes 
Ha: mean lake temperature changes with depth across different lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
lakechem.df$sampledate <- as.Date(lakechem.df$sampledate, format = "%m/%d/%Y")

#lakechem.df2 <- mutate(lakechem.df, month=month(sampledate))

lakechem.filtered <- lakechem.df %>% 
  mutate(lakechem.df, month=month(sampledate)) %>% 
  filter(month == 7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit()

unique(lakechem.filtered$month)   #verify only July was filtered for 

regress_lakes <- ggplot(lakechem.filtered, aes(x=depth, y= temperature_C))+
  geom_point(aes(color = lakename))+
  xlab("Depth (m)")+
  ylab("Temperature (C)")+
  ylim(0,35)+
  geom_smooth(method=lm, color = "black")+
  theme(legend.position = "bottom")+
  ggtitle("Temp vs. Depth Regression")
  
print(regress_lakes)

#5


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: There is a clear moderate negative correlation between temperature and depth. The distribution of points indicate this is a non-linear model, based on their curviture. An inverse logistic function would be a more appropriate estimation of their relationship, and the model's equation should be modified to account for this. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
regression_1 <- lm(lakechem.filtered$temperature_C ~ lakechem.filtered$depth)

par(mfrow = c(2,2), mar=c(4,4,4,4)) #displays all 4 graphs at once 
plot(regression_1)
par(mfrow = c(1,1)) # this cycles you through individual graphs

#depth_temp_reg <- lm(data = lakechem.filtered, temperature_C ~ depth)

summary(regression_1)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:  R2 = 0.7387 (74% of the variance in temperature can be explained by depth); df = 9726; p value of fstat < 0.05; for every 1m increase in depth, temperature falls by 1.94621 degrees C (p <0.05). 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
#create new filtered dataframe that includes all 12 explanatory variables. 
#lakechem.filtered_2 <- lakechem.df %>%     
#  mutate(lakechem.df, month=month(sampledate)) %>% 
#  filter(month == 7) %>% 
#  filter(lakeid %in% c("L", "R", "T", "E")) %>% 
#  drop_na(dissolvedOxygen, irradianceWater, irradianceDeck, lakeid)
#filtered with only year4, daynum, depth
lakechem.filtered3 <- lakechem.df %>% 
  mutate(lakechem.df, month=month(sampledate)) %>% 
  filter(month == 7) %>% 
  filter(lakeid %in% c("L", "R", "T", "E")) %>%    # filter for LTRE North Temperate Lakes 
  drop_na( lakeid, depth, lakename, year4, temperature_C) %>% 
  select(lakeid, lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit()

unique(lakechem.filtered3$lakeid)   #verification
unique(lakechem.df$lakeid)

#AIC_lakevars <- lm(data = lakechem.filtered_2, temperature_C ~ year4 + daynum + depth)   #should i use this or the next one below "AIC_lakevars2"  ?? 
#step(AIC_lakevars)

#AIC_lakevars2 <- lm(data = lakechem.filtered_2, temperature_C ~ year4 + daynum + depth + dissolvedOxygen + irradianceWater + irradianceDeck)
#step(AIC_lakevars2)

AIC_lakevars3 <- lm(data = lakechem.filtered3, temperature_C ~ year4 + daynum + depth) 
step(AIC_lakevars3)

#10
multiReg_lake <- lm(data = lakechem.filtered3, temperature_C ~ daynum + depth)
summary(multiReg_lake)

par(mfrow = c(2,2), mar=c(4,4,4,4)) #displays all 4 graphs at once 
plot(multiReg_lake)

par(mfrow = c(1,1))
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: R2 = 0.755 (76% of the variance in temperature can be explained by the explanatory variables); df = 4564; p value of fstat <0.05; for every 1m increase in depth, temperature falls by 1.99 degrees celcius. The higher R2 value is an improvement from the model that just used Depth (original r2 = 74%); the fstat holds and all of the individual explanatory variable coefficients are significant (p<0.05). 






---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#test normality of each different lake population's temperature 
#use shaprio test, 
#histograms
#bartlett test for equal variance 

#summary stats 
library(agricolae)
library(htmltools)
unique(lakechem.filtered3$lakename) # test 
laketemp.df <- lakechem.filtered3 %>% 
  group_by(lakename) %>% 
  summarise(avgtemp = mean(temperature_C))

summary(laketemp.df)

#ANOVA with aov 
lake.anova <- aov(data = lakechem.filtered3, temperature_C ~ lakename)
summary(lake.anova)

#lm 
lake.anova.lm <- lm(formula = temperature_C ~ lakename, data = lakechem.filtered3)
summary(lake.anova.lm)
par(mfrow = c(2,2), mar=c(4,4,4,4)) #displays all 4 graphs at once 
plot(lake.anova.lm)
par(mfrow = c(1,1))
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: based on the p-value of the ANOVA test being less than 0.05, there is strong evidence to support there is a significant difference in means among the lakes. This also suggests variance within each subpopulation (each lake) may not be great. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
depth_graph <- ggplot(lakechem.filtered3, aes(x= depth, y=temperature_C , color = lakename))+
  geom_point(aes(alpha= 0.5))+
  geom_smooth(method = "lm", se = FALSE)+
  ylim(0,35)+
  xlab("Depth (m)")+
  ylab("Temperature (C)")+
  theme(legend.position = "bottom")+
  ggtitle("Temp vs. Depth Regression")
  
  


print(depth_graph)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(lake.anova) 

comparison_ABC <- HSD.test(lake.anova, "lakename", group=TRUE)
comparison_ABC
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul and Peter lakes have the same mean temperatures (diff = 0.49, p value > 0.05). Both Tuesday and East Long lakes have mean temperatures statistically distinct from all of the other lakes based on examination of the p-values of the Tukey test. Both of these lakes rejected the null in comparisons with the other three lakes. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: an indepedent samples t-test or Welch's two-sample test could be used to compare the mean temperatures of these lakes. 


